---
title: "02-DataAnalysis"
author: "Danielle Ethier"
date: "22/06/2021"
output: html_document
editor_options: 
  chunk_output_type: console
---

##Monarch fall migration analysis 2021. This script is written to be stand alone from the 01-DataManip script, which is used for data download and cleaning. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Install required packages

```{r installPackages}

require(tidyverse)  
require(psych)
require(mgcv)
require(lmtest)
require(Hmisc)

detach("package:plyr", unload=TRUE) #if this was previously loaded it will cause a headache. 

#Pacakge to check model assumptions
#https://twitter.com/rfunctionaday/status/1413719791114854406
 
```

##Import data

```{r import data}

dat<-read.csv("Monarch.data.2021.csv")

dat<-dat %>% select(survey_year, mean.date, weighted.sd, quant90, doy_abund50, doy_peak1, doy_peakmean, June_temp, July_temp, August_temp, September_temp, October_temp, Summer_temp, June_tmin, July_tmin, August_tmin, September_tmin, October_tmin, Summer_temp, June_precip, July_precip, August_precip, September_precip, October_precip, Summer_precip)

```

##Peason correlation of response and predictor variables
```{r corr}

corr_response <- dat[, 2:7]
# Plot correlation matrix
pairs(corr_response)
cor(corr_response)
psych::pairs.panels(corr_response)
rcorr(as.matrix(corr_response),type="pearson")

corr_predict <- dat[, 8:24]
# Plot correlation matrix
pairs(corr_predict)
cor(corr_predict)
psych::pairs.panels(corr_predict)

cor.test(dat$Summer_temp, dat$June_temp)
cor.test(dat$Summer_temp, dat$July_temp)
cor.test(dat$Summer_temp, dat$August_temp)

#Average summer temperature correlation with June, July, and August temps. Further, mean temp is correlated with min temp. Therefore, they will be removed from the analysis
dat<-dat %>% select(-Summer_temp, -June_tmin, -July_tmin, -August_tmin, -September_tmin, -October_tmin)

```

##Explore the distribution of the response varaible to know what model distribution is more appropraite
```{r response distribution}

hist(dat$mean.date)
hist(dat$weighted.sd)
hist(dat$quant90)
hist(dat$doy_abund50)
hist(dat$doy_peak1)
hist(dat$doy_peakmean)

#These are all close enough to normally distributed to use a linear model 

```


Models developed using the methods proposed in Tredennick et al. 2021

```{r analysis Tredennick et al. 2021}

dat<-dat %>% select(survey_year, mean.date, weighted.sd, quant90, doy_abund50, doy_peak1, doy_peakmean, June_temp, July_temp, August_temp, September_temp, October_temp, June_precip, July_precip, August_precip, September_precip) %>% mutate_at(c("June_temp", "July_temp", "August_temp", "September_temp", "October_temp", "June_precip", "July_precip", "August_precip", "September_precip"), ~(scale(.) %>% as.vector))

mean<-dat %>% select(-weighted.sd, -quant90, -doy_abund50, -doy_peak1, -doy_peakmean)
pairs(mean)
cor(mean)
psych::pairs.panels(mean)
#October temp was correlated > 0.3, this is consistent with the global model above. 

#Weighted Means
mean_full<-lm(mean.date~October_temp+survey_year, data=mean)
summary(mean_full)#lm + year not significant
plot(mean_full)
drop1(mean_full, test="Chisq")

#Also look at autocorrelation in residuals
res = mean_full$res 
n = length(res) 
mod2 = lm(res[-n] ~ res[-1]) 
summary(mod2) #there is no significant temporal autocorrelation

#Weighted SD
sd<-dat %>% select(-mean.date, -quant90, -doy_abund50, -doy_peak1, -doy_peakmean)
pairs(sd)
cor(sd)
psych::pairs.panels(sd)
#June temp, August precip, > 0.3

sd_full<-lm(weighted.sd~June_temp+August_precip+survey_year, data=sd)
summary(sd_full)
plot(sd_full)
drop1(sd_full, test="Chisq") #Index significant

#90th percentile
p90<-dat %>% select(-mean.date, -weighted.sd, -doy_abund50, -doy_peak1, -doy_peakmean)
pairs(p90)
cor(p90)
psych::pairs.panels(p90)
#September temp was correlated > 0.3, this is consistent with the global model

p90_full<-lm(quant90~September_temp+survey_year, data=p90)
summary(p90_full) #September temp significant
plot(p90_full)
drop1(p90_full, test="Chisq")

#look at autocorrelation in residuals
res = p90_full$res 
n = length(res) 
mod2 = lm(res[-n] ~ res[-1]) 
summary(mod2) #there is no significant temporal autocorrelations

#plot significant results (quant90)
ggplot(p90, aes (x=September_temp, y=quant90)) +
  geom_point() +
  geom_smooth(method ="lm", colour="black", size=0.5) + 
  ylab("Late season passage (90th quantile)") +
  xlab("September temperature (C)") +
  theme(text=element_text(size=20))+
  theme_classic()

#50% abund passage 
abund<-dat %>% select(-mean.date, -weighted.sd, -quant90, -doy_peak1, -doy_peakmean)
pairs(abund)
cor(abund)
psych::pairs.panels(abund)
#October temp was correlated > 0.3

abund_m<-lm(doy_abund50~October_temp+survey_year, data=abund)
summary(abund_m) #October temp not significant
drop1(abund_m, test="Chisq")

#first peak passage 
peak<-dat %>% select(-mean.date, -weighted.sd, -quant90, -doy_abund50, -doy_peakmean)
pairs(peak)
cor(peak)
psych::pairs.panels(peak)
#October temp was correlated > 0.3

peak_m<-lm(doy_peak1~October_temp+survey_year, data=peak)
summary(peak_m) #October temp not significant
drop1(peak_m, test="Chisq")

#mean peak passage 
meanpeak<-dat %>% select(-mean.date, -weighted.sd, -quant90, -doy_abund50, -doy_peak1)
pairs(meanpeak)
cor(meanpeak)
psych::pairs.panels(meanpeak)
#October temp was correlated > 0.3

peak_m2<-lm(doy_peakmean~October_temp+survey_year, data=meanpeak)
summary(peak_m2) #October temp not significant, but close at 0.03 
drop1(peak_m2, test="Chisq")


```

Plot response over time

```{r plot}

#plot mean.data with time
ggplot(dat, aes (x=survey_year, y=mean.date)) +
  geom_point() +
  geom_smooth(method ="lm", colour="black", size=0.5) + 
  ylab("Mean passage (weighted)") +
  xlab("Year") +
  theme(text=element_text(size=20))+
  theme_classic()

#plot sd. weighted with time
ggplot(dat, aes (x=survey_year, y=weighted.sd)) +
  geom_point() +
  geom_smooth(method ="lm", colour="black", size=0.5) + 
  ylab("Standard deviation (weighted)") +
  xlab("Year") +
  theme(text=element_text(size=20))+
  theme_classic()

#plot mean.passage 50% with time
ggplot(dat, aes (x=survey_year, y=doy_abund50)) +
  geom_point() +
  geom_smooth(method ="lm", colour="black", size=0.5) + 
  ylab("Midpoint passage (50 percent)") +
  xlab("Year") +
  theme(text=element_text(size=20))+
  theme_classic()

#plot doy_peak1 with time
ggplot(dat, aes (x=survey_year, y=doy_peak1)) +
  geom_point() +
  geom_smooth(method ="lm", colour="black", size=0.5) + 
  ylab("First peak day") +
  xlab("Year") +
  theme(text=element_text(size=20))+
  theme_classic()

#plot doy_peakmean with time
ggplot(dat, aes (x=survey_year, y=doy_peakmean)) +
  geom_point() +
  geom_smooth(method ="lm", colour="black", size=0.5) + 
  ylab("Average peak day") +
  xlab("Year") +
  theme(text=element_text(size=20))+
  theme_classic()

#plot quant90 with time
ggplot(dat, aes (x=survey_year, y=quant90)) +
  geom_point() +
  geom_smooth(method ="lm", colour="black", size=0.5) + 
  ylab("Late passage (90th percentile)") +
  xlab("Year") +
  theme(text=element_text(size=20))+
  theme_classic()

```